---
title: "Psychiatric Predictors of Violent Traumatic Brain Injury"
subtitle: "<b>Results from the TBIMS Study</b><br> In this study, we estimated the association between psychiatric illnesses/indices and TBI aetiology using pubically available data from the the Traumatic Brain Injury Model Systems National Database."

format:
  lumo-html: 
    #logo:  #optional: add your logo to the doc header
    primary-color: "#b5cef7"
    github-repo: "https://github.com/mayaogonah" 
    self-contained: true
    is-particlejs-enabled: true  # optional: display interactive particles in the doc header
    #bg-image:  # optional: add a background image
    code-fold: show

execute:
  eval: false
  echo: true
author: Maya Ogonah (Columbia University)
email: mgo2120@columbia.edu
date: today
---
## Libraries and data

### Packages

First load all necessary packages.

```{r}
#install.packages("remotes")
#remotes::install_git("https://codeberg.org/hrbrmstr/hrbrthemes.git")

```

```{r, warning=FALSE, message=FALSE}
if (!require("pacman")) install.packages("pacman")

library(pacman)
pacman::p_load(
  remotes, # used to intstall R packages from remote/local repositories, eg, GitHub
  styler, # formats code automatically
  readxl, # read Excel files
  DT, # interface for DataTables
  kableExtra, # scrollable tables
  tidyverse, # data wrangling 
  labelled, # handling labelled variables
  devtools, # package development
  modelsummary, # create tables and plots for descriptive statistics
  estimatr, # calculates difference in means with uncertainty estimates (modelsummary)
  ggplot2, # data visualisations and plots
  ggridges, # ggplot2 extension for ridgeline plots
  cowplot, # ggplot add on 
  MetBrewer, # <1>
  hrbrthemes, # theme components for ggplot2
  tinytable, # editing/formatting tables
  knitr, # report generation
  gt, # presentation-level tables
  tibble, # utilities for handling tibbles 
  naniar, # missing data 
  VIM, # visualise missing data 
  faraway, #ilogit() function
  sjPlot, # plotting ORs
  gtsummary, # presentation of regression analyses
  rms, # assessing the linearity assumption with restricted cubic splines
  reshape, # for the melt() function
  purrr, # for functions and vectors
  haven, # missing data MCAR
  finalfit, # missing data MCAR
  mice, # multiple imputation
  miceadds, # multiple imputation
  RColorBrewer # palettes proposed by RColorBrewer for heatmap
)

# download custon functions from the "Introduction to Regression Methods for Public Health Using R" by Ramzi W Nahhas
source("Functions_rmph.R")

```

1.  This is one of my favourite packages, [MetBrewer](https://github.com/BlakeRMills/MetBrewer/blob/main/README.md). It allows you to utilise palettes inspired by works at the Metropolitan Museum of Art in New York. This Quarto file uses the Klimt palette.

### Dataviz

Create a datavisualisation theme.

```{r}
scale_color_brand <- function() {
  custom_palette <- MetBrewer::met.brewer(name = "Klimt", n=20, type="continuous") # <2>
  scale_color_manual(values = custom_palette)
}

scale_fill_brand <- function(labels = NULL) {
  custom_palette <- MetBrewer::met.brewer(name = "Klimt", n=20, type="continuous")
  if (is.null(labels)) {
    scale_fill_manual(values = custom_palette)
  } else {
    scale_fill_manual(values = custom_palette, labels = labels)
  }
}

palette <- MetBrewer::met.brewer(name = "Klimt", n=20, type="continuous")

theme_brand <- function() {
  theme_ipsum() + # <3>
    theme(
      plot.title = element_text(color = "black", size = 14, face = "bold"),
      plot.subtitle = element_text(margin = margin(b = 20)),
      axis.text.x = element_text(size = 7),
      axis.text.y = element_text(size = 7),
      axis.title.x = element_text(hjust = 0.5),  # center x-axis label
      axis.title.y = element_text(hjust = 0.5),  # center y-axis label
      plot.background = element_rect(fill = "gray97", color = NA),
      plot.caption = element_text(hjust = 0)
    )
}
```

2.  The n function allows you to extend the number of values so the colour palette can still be used on graphs with numerous categorical values.
3.  You can change the font using the hrbrthemes package. I don't have the other fonts installed, and its a bit of a pain to install them, hence why I used theme_ipsum.


### Import data
Then, import the the publicly available de-identified data from [TBIMS](https://www.tbindsc.org/default.aspx), and save the .csv file to a R studio project (no need to setwd).

```{r, warning=FALSE, message=FALSE}
data <- read_csv("data/input/TBIMSForm1_Public_20250812.csv")
data_long <- read_csv("data/input/TBIMSForm2_Public_20250812.csv")
```


# Data linkage and wrangling

## Data linkage

Now lets link Form 1 and Form 2 using union/vertical joins. Data is in the long format.

```{r}
# create baseline rows from baseline data
baseline_long <- data %>%
  transmute(
    Mod1Id,
    Mod2Id = NA,
    FollowUpPeriod = 0,
    AGENoPHIF = AGENoPHI,
    EduYearsF = EduYears,
    EDUCATIONF = EDUCATION,
    EMPLOYMENTF = EMPLOYMENT,
    EarnF = Earn,
    RURALF = RURALadm,
    CombinedDRSF = DRSa,
    FIMTOTF = FIMTOTA,
    FIMCOGF = FIMCOGA,
    DRINKCatF = DRINKCat,
    PROBLEMUseF = PROBLEMUse,
    DrugsF = Drugs,
    SuicideF = Suicide,
    TBIInjuryF = TBIInjury,
    TBIRepInjuryF = TBIRepInjury,
    DeathCause1F = DeathCause1,
    DeathCause2F = DeathCause2,
    DeathECodeF = DeathECode
  ) # `transmute()` creates NEW columns and DROPS all other columns (unlike `mutate()` which keeps everything)

# create follow-up rows from follow-up data
followup_long <- data_long %>%
  select(Mod1Id, Mod2Id, FollowUpPeriod, AGENoPHIF, EduYearsF, EDUCATIONF, EarnF, EMPLOYMENTF, CombinedDRSF, FIMTOTF, FIMCOGF, GOSTotalF, DRINKCatF, DrugsF, PROBLEMUseF, SuicideF, TBIInjuryF, TBIRepInjuryF, PHQPleasureF, PHQDownF, PHQ9TOTF, GADNervousF, GADCntrlWryF, GAD7TOTF, DeathCause1F, DeathCause2F, DAYStoDEATHF, DeathECodeF, HHIncomeF)

# combine baseline and follow-up (union/vertical joins)
longitudinal_data <- bind_rows(baseline_long, followup_long) %>%
  arrange(Mod1Id, FollowUpPeriod)

# add time-invariant demographics from baseline data
longitudinal_complete <- longitudinal_data %>%
  left_join(
    data %>%
      select(Mod1Id, SexF, RaceIndF, RaceBlkF, RacePIF, RaceAsnF, RaceWhtF,
             EthnicityF, CountryBirthF, INJYEAR, DEATHYEAR, 
             Cause, GCS, SCI, CC_Hypertension, SpEd, SuicidePrior, PsyHosp, MntlEver),
    by = "Mod1Id"
  )
```

Check the unique number of individuals in the linked data.

```{r}
longitudinal_complete %>%
  distinct(Mod1Id) %>%
  nrow()

```


```{r}
table(longitudinal_complete$EMPLOYMENTF)
```


## Data cleaning

Clean up the data. Ensure categorical variables are coded as factors as this makes it much easier to use in statistical models and graphs. Age, year, and time-to-event variables are coded as numeric. "Unknown" or missing data is coded as NA. Invalid values are also coded as NA.

### Demographics

```{r}
# age (AGENoPHIF)
longitudinal_complete <- longitudinal_complete %>%
  mutate(AGENoPHIF_C = ifelse(AGENoPHIF < 16, NA, AGENoPHIF)) %>% # those with an incorrect age coded as NA
  mutate(AGENoPHIF_C = ifelse(AGENoPHIF_C == 999, NA, AGENoPHIF_C)) %>% # a cleaned version where missingness is denoted by NA
  mutate(AGENoPHIF_C = ifelse(AGENoPHIF_C == 777, 89, AGENoPHIF_C)) # windorise those aged over 89 to 89

# fill in missing age using baseline age + follow-up time
longitudinal_complete <- longitudinal_complete %>%
  group_by(Mod1Id) %>%
  mutate(
    baseline_age = first(AGENoPHIF_C[FollowUpPeriod == 0], default = NA_real_),
    AGENoPHIF_C = if_else(is.na(AGENoPHIF_C),
                        baseline_age + FollowUpPeriod,
                        AGENoPHIF_C)
  ) %>%
  select(-baseline_age) %>%
  ungroup() %>%
  mutate(AGENoPHIF_C = as.numeric(AGENoPHIF_C))

# sex (SexF)
longitudinal_complete <- longitudinal_complete %>%
  mutate(SexF = as.factor(SexF)) %>%
  mutate(SexF_C = ifelse(SexF == 99 | is.na(SexF), NA, SexF)) %>%
  mutate(SexF_C = as.factor(SexF_C))

# race (RaceCat)
# the original `Race` variable (added in 1989) was archived in 2023 and is therefore not available in the publicly available dataset. During data cleaning I created a categorical variable for race using `RaceWhtF`, `RaceBlkF`, `RacePIF`, `RaceIndF` and `RaceAsnF` and those who identified with more than 1 race.
## categorise race
longitudinal_complete <- longitudinal_complete %>%
  mutate(
    RaceIndF_C = ifelse(RaceIndF > 1, NA, RaceIndF),
    RaceBlkF_C = ifelse(RaceBlkF > 1, NA, RaceBlkF),
    RacePIF_C = ifelse(RacePIF > 1, NA, RacePIF),
    RaceAsnF_C = ifelse(RaceAsnF > 1, NA, RaceAsnF),
    RaceWhtF_C = ifelse(RaceWhtF > 1, NA, RaceWhtF)
  ) %>%
  mutate(
    RaceCount =  RaceIndF_C + RaceBlkF_C + RacePIF_C + RaceAsnF_C + RaceWhtF_C,
    RaceCat = case_when(
      RaceCount == 1 & RaceWhtF_C == 1 ~ 1, # white
      RaceCount == 1 & RaceBlkF_C == 1 ~ 2, # black
      RaceCount == 1 & RacePIF_C == 1 ~ 3, # PIF
      RaceCount == 1 & RaceAsnF_C == 1 ~ 4, # Asian 
      RaceCount == 1 & RaceIndF_C == 1 ~ 5, # AIAN
      RaceCount > 1 ~ 6,  #mixed race category
      is.na(RaceBlkF_C) & is.na(RacePIF_C) & is.na(RaceAsnF_C) & is.na(RaceIndF_C) & !is.na(RaceWhtF_C) ~ 1, # white
      !is.na(RaceBlkF_C) & is.na(RaceWhtF_C) & is.na(RacePIF_C) & is.na(RaceAsnF_C) & is.na(RaceIndF_C) ~ 2, # black
    !is.na(RacePIF_C) & is.na(RaceWhtF_C) & is.na(RaceBlkF_C) & is.na(RaceAsnF_C) & is.na(RaceIndF_C) ~ 3, # NHPI
    !is.na(RaceAsnF_C) & is.na(RaceWhtF_C) & is.na(RaceBlkF_C) & is.na(RacePIF_C) & is.na(RaceIndF_C) ~ 4, # Asian 
    !is.na(RaceIndF_C) & is.na(RaceWhtF_C) & is.na(RaceBlkF_C) & is.na(RacePIF_C) & is.na(RaceAsnF_C) ~ 5, # AIAN
    )
  ) %>%
  mutate(Race1 = case_when(
    RaceCat == 1 ~ 1, # white
    RaceCat == 2 ~ 2, # black
    RaceCat == 4 ~ 3, # asian
    RaceCat == 3 ~ 4, # NHPI
    RaceCat == 5 ~ 5, # AIAN
    RaceCat == 6 ~ 6 # mixed race
  )) %>%
  mutate(RaceCat = as.factor(RaceCat)) %>%
  mutate(Race1 = as.factor(Race1))

# ethnicity (EthnicityF)
longitudinal_complete <- longitudinal_complete %>%
  mutate(EthnicityF = as.factor(EthnicityF)) %>%
  mutate(EthnicityF_C= ifelse(EthnicityF %in% c(66, 77, 99) | is.na(EthnicityF), NA, EthnicityF)) %>%
  mutate(EthnicityF_C = as.factor(EthnicityF_C))

# race_ethnicity
longitudinal_complete <- longitudinal_complete %>%
  mutate(
    race_ethnicity = case_when(
      # hispanic 
      EthnicityF_C == 2 & Race1 == 1 ~ "Hispanic White",
      EthnicityF_C == 2 & Race1 == 2 ~ "Hispanic Black",
      EthnicityF_C == 2 & Race1 == 3 ~ "Hispanic Asian",
      EthnicityF_C == 2 & Race1 == 4 ~ "Hispanic NHPI",
      EthnicityF_C == 2 & Race1 == 5 ~ "Hispanic AIAN",
      EthnicityF_C == 2 & Race1 == 6 ~ "Hispanic Mixed Race",
      EthnicityF_C == 2 & is.na(Race1) ~ "Hispanic (Race Not Reported)",

      # non-hispanic
      EthnicityF_C == 1 & Race1 == 1 ~ "Non-Hispanic White",
      EthnicityF_C == 1 & Race1 == 2 ~ "Non-Hispanic Black",
      EthnicityF_C == 1 & Race1 == 3 ~ "Non-Hispanic Asian",
      EthnicityF_C == 1 & Race1 == 4 ~ "Non-Hispanic NHPI",
      EthnicityF_C == 1 & Race1 == 5 ~ "Non-Hispanic AIAN",
      EthnicityF_C == 1 & Race1 == 6 ~ "Non-Hispanic Mixed Race",
      EthnicityF_C == 1 & is.na(Race1) ~ "Non-Hispanic (Race Not Reported)",

      # race only
      is.na(EthnicityF_C) & Race1 == 1 ~ "White (Ethnicity Not Reported)",
      is.na(EthnicityF_C) & Race1 == 2 ~ "Black (Ethnicity Not Reported)",
      is.na(EthnicityF_C) & Race1 == 3 ~ "Asian (Ethnicity Not Reported)",
      is.na(EthnicityF_C) & Race1 == 4 ~ "NHPI (Ethnicity Not Reported)",
      is.na(EthnicityF_C) & Race1 == 5 ~ "AIAN (Ethnicity Not Reported)",
      is.na(EthnicityF_C) & Race1 == 6 ~ "Mixed Race (Ethnicity Not Reported)",

      # neither race or ethnicity reported
      TRUE ~ NA_character_
    )
  ) %>%
  
  # NIH reporting categories
  mutate(
    race_ethnicity_C = case_when(
      # Hispanic ALWAYS overrides race
      str_starts(race_ethnicity, "Hispanic") ~ "Hispanic or Latino",
      
      # to reduce missingness → Non-Hispanic other is a category
      str_detect(race_ethnicity, "Non-Hispanic \\(Race Not Reported\\)") ~ "Non-Hispanic Other",

      # race-based NIH categories
      str_detect(race_ethnicity, "Asian") | str_detect(race_ethnicity, "NHPI") ~ "Asian / NHPI",
      str_detect(race_ethnicity, "Black") ~ "Black or AA",
      str_detect(race_ethnicity, "White") ~ "White",
      str_detect(race_ethnicity, "AIAN") ~ "AIAN",

      # If mixed and non-Hispanic → NIH would classify as "More than one race"
      str_detect(race_ethnicity, "Mixed") ~ "More than one race",

      TRUE ~ "Unknown"
    )
  ) %>%

  mutate(race_ethnicity_C = case_when(
    race_ethnicity_C == "White" ~ 1,
    race_ethnicity_C == "Black or AA" ~ 2,
    race_ethnicity_C == "Asian / NHPI" ~ 3,
    race_ethnicity_C == "Hispanic or Latino" ~ 4,
    race_ethnicity_C == "AIAN" ~ 5,
    race_ethnicity_C == "Non-Hispanic Other" ~ 6,
    race_ethnicity_C == "More than one race" ~ 7,
    race_ethnicity_C == "Unknown" ~ NA
  ),
  race_ethnicity_C = as.factor(race_ethnicity_C))

# rural (RURALF)
longitudinal_complete <- longitudinal_complete %>%
  mutate(RURALF = as.factor(RURALF))

# country of birth (CountryBirthF)
longitudinal_complete <- longitudinal_complete %>%
  mutate(CountryBirthF = as.factor(CountryBirthF)) %>%
  mutate(CountryBirthF_C= ifelse(CountryBirthF %in% c(77, 99) | is.na(CountryBirthF), NA, CountryBirthF)) %>%
  mutate(CountryBirthF_C = as.factor(CountryBirthF_C))

# education (EduYearsF)
## categorise those with less than 10 years of school as those "still in school"
longitudinal_complete <- longitudinal_complete %>%
  mutate(EduYears10 = case_when(
    EduYearsF <= 10 ~ 1,
    EduYearsF == 11 ~ 2,
    EduYearsF == 12 ~ 3,
    EduYearsF == 13 ~ 4,
    EduYearsF == 14 ~ 5,
    EduYearsF == 15 ~ 6,
    EduYearsF == 16 ~ 7,
    EduYearsF == 17 ~ 8,
    EduYearsF == 18 ~ 9,
    EduYearsF == 19 ~ 10,
    EduYearsF == 20 ~ 11,
    EduYearsF == 21 ~ 12,
    EduYearsF == 666 ~ 666,
    EduYearsF == 999 ~ 999
  )) %>%
  mutate(EduYears10_C = if_else(EduYears10 %in% c(666, 999) | is.na(EduYears10), NA, EduYears10)) %>%
  mutate(EDUCATION_C = case_when(
           EDUCATIONF == 1 | EDUCATIONF == 2 ~ 1, # no high school degree
           EDUCATIONF == 2.5 | EDUCATIONF == 3 | EDUCATIONF == 3.5 ~ 2, # HS/GED
           EDUCATIONF == 5 ~ 3, # some college
           EDUCATIONF == 6 ~ 4, # associated degree
           EDUCATIONF == 7 ~ 5, # bachelors
           EDUCATIONF == 8 | EDUCATIONF == 9 ~ 6, # graduate degree
           EDUCATIONF == 4 | EDUCATIONF == 21 ~ NA, # other (due to the small number of individuals with this category they were bundled with unknown)
           EDUCATIONF == 999 ~ NA
           )) %>%
  mutate(EduYears10_C = as.factor(EduYears10_C)) %>%
  mutate(EDUCATIONF = as.factor(EDUCATIONF)) %>%
  mutate(EDUCATION_C = as.factor(EDUCATION_C))

# income (EarnF)
longitudinal_complete <- longitudinal_complete %>%
  mutate(EarnF = as.factor(EarnF)) %>%
  mutate(EarnF_C = if_else(EarnF %in% c(666, 777, 999) | is.na(EarnF), NA, EarnF)) %>%
  mutate(EarnF_C = as.factor(EarnF_C))

# household income (HHIncomeF)
longitudinal_complete <- longitudinal_complete %>%
  mutate(HHIncomeF = as.factor(HHIncomeF)) %>%
  mutate(HHIncomeF_C = if_else(HHIncomeF %in% c(66, 77, 99) | is.na(HHIncomeF), NA, HHIncomeF)) %>%
  mutate(HHIncomeF_C = as.factor(HHIncomeF_C))

# special education (SpEd)
## those with NA are assigned the value of "99 = Unknown"
longitudinal_complete <- longitudinal_complete %>%
  mutate(SpEd = as.factor(SpEd)) %>%
  mutate(SpEd_C= ifelse(SpEd %in% c(66, 77, 99) | is.na(SpEd), NA, SpEd)) %>%
  mutate(SpEd_C = as.factor(SpEd_C))

# employment (EMPLOYMENT)
## those with NA are assigned the value of "999 = Unknown"
longitudinal_complete <- longitudinal_complete %>%
  mutate(EMPLOYMENTF_C = ifelse(EMPLOYMENTF == 999 | is.na(EMPLOYMENTF), NA, EMPLOYMENTF)) %>%
  mutate(EMPLOY1 = case_when(
    EMPLOYMENTF_C == 2 | EMPLOYMENTF_C == 3  ~ 2, # student (part or full time)
    EMPLOYMENTF_C == 5 ~ 1, # competitively employed
    EMPLOYMENTF_C == 9 ~ 3, # retired
    EMPLOYMENTF_C == 4 | EMPLOYMENTF_C == 7 | EMPLOYMENTF_C == 11 | EMPLOYMENTF_C == 12 | EMPLOYMENTF_C == 8 ~ 4, # other
    EMPLOYMENTF_C == 10 ~ 5 # unemployment
  )) %>%
  mutate(EMPLOYMENTF_C = as.factor(EMPLOYMENTF_C)) %>% 
  mutate(EMPLOYMENTF = as.factor(EMPLOYMENTF)) %>%
  mutate(EMPLOY1 = as.factor(EMPLOY1))

```

### Injury characteristics

```{r}
# injury year
longitudinal_complete <- longitudinal_complete %>%
  mutate(INJYEAR = as.numeric(INJYEAR))

# cause of TBI (CauseCat)
## combine the cause of injury categories into: vehicular, violence, sports, falls, other, and unknown
longitudinal_complete <- longitudinal_complete %>%
  mutate(Cause = as.factor(Cause)) %>%
  mutate(CauseCat = case_when(
    (Cause == 1 | Cause == 2 | Cause == 3 | Cause == 4 | Cause == 5) ~ 1, # vehicular
    Cause == 10 ~ 2, #gunshot wound
    Cause == 11 ~ 3, # assault with blunt instrument
    Cause == 12 ~ 4, # other violence
    (Cause == 13 | Cause == 14 | Cause == 15 | Cause == 16 | Cause == 17 | Cause == 18) ~ 5, # sports
    (Cause == 19) ~ 6, # falls
    (Cause == 20 | Cause == 21 | Cause == 22) ~ 7, # other
    (Cause == 999) ~ NA, TRUE ~ NA
  )) %>%
  mutate(CauseCat = ifelse(CauseCat == 999 | is.na(CauseCat), NA, CauseCat)) %>%
  mutate(Aetiology = case_when(
    CauseCat == 2 | CauseCat == 3 | CauseCat == 4 ~ 1,
    CauseCat ==  1 | CauseCat ==  5 | CauseCat ==  6 | CauseCat ==  7 ~ 0,
    TRUE ~ NA)
  ) %>%
  mutate(CauseCat = as.factor(CauseCat)) %>%
  mutate(Aetiology = as.factor(Aetiology)) 

# GCS scores includ. sedation category (GCSCatSed)
## those with NA are assigned the value of "999 = Unknown"
## recode the GCS variable into a categorical variable based on the criteria: mild (scores ≥ 13), moderate (scores 9–12), severe (scores 3–8), sedation (GCS = 77), intubation (GCS = 88), and missing (scores = 999)
longitudinal_complete <- longitudinal_complete %>%
  mutate(GCSCatSed = case_when(
    GCS < 9 ~ 1, # severe
    (GCS >= 9 & GCS < 13) ~ 2, # moderate
    (GCS >= 13 & GCS < 77) ~ 3, # mild
    GCS == 77 ~ 77, # sedated
    GCS == 88 ~ 88, # intubated
    GCS == 999 ~ 999,
    TRUE ~ GCS
  )) %>%
  mutate(GCSCatSed = ifelse(GCSCatSed == 999 | is.na(GCSCatSed), NA, GCSCatSed)) %>%
  mutate(GCSCatSed = as.factor(GCSCatSed)) %>%
  mutate(GCS = as.factor(GCS))

# spinal cord injury
longitudinal_complete <- longitudinal_complete %>%
  mutate(SCI_C = ifelse(SCI == 99 | is.na(SCI), NA, SCI)) %>%
  mutate(SCI_C = as.factor(SCI_C))

# hypertension
longitudinal_complete <- longitudinal_complete %>%
  mutate(CC_Hypertension = as.factor(CC_Hypertension)) %>%
  mutate(CC_Hypertension_C = ifelse(CC_Hypertension == 99 | is.na(CC_Hypertension), NA, CC_Hypertension)) %>%
  mutate(CC_Hypertension_C = as.factor(CC_Hypertension_C))
```

### History / repeated TBI

```{r}
# TBI history
longitudinal_complete <- longitudinal_complete %>%
  mutate(TBIInjuryF = as.factor(TBIInjuryF)) %>%
  mutate(TBIInjuryF_C = if_else(TBIInjuryF %in% c(66, 77, 99) | is.na(TBIInjuryF), NA, TBIInjuryF)) %>%
  mutate(TBIInjuryF_C = as.factor(TBIInjuryF_C))

# TBI repeated TBI
longitudinal_complete <- longitudinal_complete %>%
  mutate(TBIRepInjuryF = as.factor(TBIRepInjuryF)) %>%
  mutate(TBIRepInjuryF_C = if_else(TBIRepInjuryF %in% c(66, 77, 99) | is.na(TBIRepInjuryF), NA, TBIRepInjuryF)) %>%
  mutate(TBIRepInjuryF_C = as.factor(TBIRepInjuryF_C))
```

### Disability

```{r}
# disability rating scale (DRSF)
longitudinal_complete <- longitudinal_complete %>%
  mutate(DRSF_C = ifelse(CombinedDRSF %in% c(777, 999) | is.na(CombinedDRSF), NA, CombinedDRSF)) # a cleaned version where missingness is denoted by NA

# FIM total (FIMTOTF)
longitudinal_complete <- longitudinal_complete %>%
  mutate(FIMTOTF_C = ifelse(FIMTOTF == 9999 | is.na(FIMTOTF), NA, FIMTOTF)) # a cleaned version where missingness is denoted by NA
# FIM cognitive (FIMCOGF)
longitudinal_complete <- longitudinal_complete %>%
  mutate(FIMCOGF_C = ifelse(FIMCOGF == 9999 | is.na(FIMCOGF), NA, FIMCOGF)) # a cleaned version where missingness is denoted by NA

# GOS-E (GOSTotalF)
longitudinal_complete <- longitudinal_complete %>%
  mutate(GOSTotalF_C = ifelse(GOSTotalF %in% c(66, 99) | is.na(GOSTotalF), NA, GOSTotalF)) 
```

### Cause of death

```{r}
# age at death
longitudinal_complete <- longitudinal_complete %>%
  mutate(
    DAYStoDEATHF_C = ifelse(DAYStoDEATHF == 999999, NA, DAYStoDEATHF),
    AgeDeath = case_when(
      is.na(DAYStoDEATHF_C) ~ NA,
      TRUE ~ (DAYStoDEATHF_C/365.25)))
## imputed age at death when the DAYStoDEATHF variable is missing
longitudinal_complete <- longitudinal_complete %>%
  group_by(Mod1Id) %>%
  mutate(
    AgeNoPHI0 = AGENoPHIF_C[FollowUpPeriod == 0][1],
    AgeDeath_C = case_when(
      !is.na(AgeDeath) ~ AgeDeath,
      !is.na(DEATHYEAR) & is.na(AgeDeath) ~ (AgeNoPHI0 + (DEATHYEAR - INJYEAR))
    )
  ) %>%
  ungroup() %>%
  select(-AgeNoPHI0) %>%
  mutate(AgeDeath_C = as.numeric(AgeDeath_C))
## still cannot compute an age at death for those without an age at baseline or DAYStoDEATHF (n = 7)

## death by assault
longitudinal_complete <- longitudinal_complete %>%
  mutate(DeathAssault = case_when(
    DeathECodeF %in% c("E9060", "E960.0", "E960.1", "E961", "E962", "E962.0", "E962.1", "E962.2", "E962.9", "E963", "E964", "E965", "E965.0", "E965.1", "E965.2", "E965.3", "E965.4", "E965.5", "E965.6", "E965.7", "E965.8", "E965.9", "E966", "E967", "E967.0", "E967.1", "E967.2", "E967.3", "E967.4", "E967.5", "E967.6", "E967.7", "E967.8", "E967.9", "E968", "E968.0", "E968.1", "E968.2", "E968.3", "E968.4", "E968.5", "E968.6", "E968.7", "E968.8", "E968.9", "E969") ~ 1,
    TRUE ~ 0
  ),
  DeathAssault = as.factor(DeathAssault)) 

## death by suicide
longitudinal_complete <- longitudinal_complete %>%
  mutate(DeathSuicide = case_when(
    DeathECodeF %in% c("E950", "E950.0", "E950.1", "E950.2", "E950.3", "E950.4", "E950.5", "E950.6", "E950.7", "E950.8", "E950.9", "E951", "E951.0", "E951.1", "E951.8", "E952", "E952.0", "E952.1", "E952.8", "E952.9", "E953", "E953.1", "E953.8", "E953.9", "E954", "E955", "E955.0", "E955.1", "E955.2", "E955.3", "E955.4", "E955.5", "E955.6", "E955.7", "E955.9", "E956", "E957", "E957.0", "E957.1", "E957.2", "E957.9", "E958", "E958.0", "E958.1", "E958.2", "E958.3", "E958.4", "E958.5", "E958.6", "E958.7", "E958.8", "E958.9", "E959", "E980", "E980.0", "E980.1", "E980.2", "E980.3", "E980.4", "E980.5", "E980.6", "E980.7", "E980.8", "E980.9", "E981", "E981.0", "E981.1", "E981.8", "E982", "E982.0", "E982.1", "E982.8", "E982.9", "E983", "E983.0", "E983.1", "E983.8", "E983.9", "E984", "E985", "E985.0", "E985.1", "E985.2", "E985.3", "E985.4", "E985.5", "E985.6", "E985.7", "E986", "E987", "E987.0", "E987.1", "E987.2", "E987.9", "E988", "E988.0", "E988.1", "E988.2", "E988.3", "E988.4", "E988.5", "E988.6", "E988.7", "E988.8", "E988.9", "E989") ~ 1,
    TRUE ~ 0
  ),
  DeathSuicide = as.factor(DeathSuicide))

## death by overdose
longitudinal_complete <- longitudinal_complete %>%
  mutate(DeathOverdose = case_when(
    DeathCause1F %in% c("291", "291.0", "291.1", "291.2", "291.3", "291.4", "291.5", "291.8", "291.81", "291.82", "291.89", "291.9", "292", "292.0", "292.1", "292.11", "292.12", "292.2", "292.8", "292.81", "292.82", "292.83", "292.84", "292.85", "292.89", "292.9", "303", "303.0", "303.00", "303.01", "303.02", "303.03", "303.9", "303.90", "303.91", "303.92", "303.93", "304", "304.0", "304.00", "304.01", "304.02", "304.03", "304.1", "304.10", "304.11", "304.12", "304.13", "304.2", "304.20", "304.21", "304.22", "304.23", "304.3", "304.30", "304.31", "304.32", "304.33", "304.4", "304.40", "304.41", "304.42", "304.43", "304.5", "304.50", "304.51", "304.52", "304.53", "304.6", "304.60", "304.61", "304.62", "304.63", "304.7", "304.70", "304.71", "304.72", "304.73", "304.8", "304.80", "304.81", "304.82", "304.83", "304.9", "304.90", "304.91", "304.92", "304.93", "305", "305.0", "305.00", "305.01", "305.02", "305.03", "305.2", "305.20", "305.21", "305.22", "305.23", "305.3", "305.30", "305.31", "305.32", "303.33", "305.4", "305.40", "305.41", "305.42", "305.43", "305.5", "305.50", "305.51", "305.52", "305.53", "305.6", "305.60", "305.61", "305.62", "305.63", "305.7", "305.70", "305.71", "305.72", "305.73", "305.8", "305.80", "305.81", "305.82", "305.83", "305.9", "305.90", "305.91", "305.92", "305.93") ~ 1,
    DeathCause2F %in% c("291", "291.0", "291.1", "291.2", "291.3", "291.4", "291.5", "291.8", "291.81", "291.82", "291.89", "291.9", "292", "292.0", "292.1", "292.11", "292.12", "292.2", "292.8", "292.81", "292.82", "292.83", "292.84", "292.85", "292.89", "292.9", "303", "303.0", "303.00", "303.01", "303.02", "303.03", "303.9", "303.90", "303.91", "303.92", "303.93", "304", "304.0", "304.00", "304.01", "304.02", "304.03", "304.1", "304.10", "304.11", "304.12", "304.13", "304.2", "304.20", "304.21", "304.22", "304.23", "304.3", "304.30", "304.31", "304.32", "304.33", "304.4", "304.40", "304.41", "304.42", "304.43", "304.5", "304.50", "304.51", "304.52", "304.53", "304.6", "304.60", "304.61", "304.62", "304.63", "304.7", "304.70", "304.71", "304.72", "304.73", "304.8", "304.80", "304.81", "304.82", "304.83", "304.9", "304.90", "304.91", "304.92", "304.93", "305", "305.0", "305.00", "305.01", "305.02", "305.03", "305.2", "305.20", "305.21", "305.22", "305.23", "305.3", "305.30", "305.31", "305.32", "303.33", "305.4", "305.40", "305.41", "305.42", "305.43", "305.5", "305.50", "305.51", "305.52", "305.53", "305.6", "305.60", "305.61", "305.62", "305.63", "305.7", "305.70", "305.71", "305.72", "305.73", "305.8", "305.80", "305.81", "305.82", "305.83", "305.9", "305.90", "305.91", "305.92", "305.93") ~ 2,
    TRUE ~ 0
  ),
  DeathOverdose = as.factor(DeathOverdose)) 
```

### Psychiatric illness and indices

```{r}
# pre-TBI history of psychiatric hospitalisation (PsyHosp)
longitudinal_complete <- longitudinal_complete %>%
  mutate(PsyHosp = as.factor(PsyHosp)) %>%
  mutate(PsyHosp_C = ifelse(PsyHosp %in% c(66, 77, 88, 99) | is.na(PsyHosp), NA, PsyHosp)) %>%
  mutate(PsyHosp_C = as.factor(PsyHosp_C))

# pre-TBI history of mental health treatment (MntlEver)
longitudinal_complete <- longitudinal_complete %>%
  mutate(MntlEver = as.factor(MntlEver)) %>%
  mutate(MntlEver_C = ifelse(MntlEver %in% c(66, 77, 99) | is.na(MntlEver), NA, MntlEver)) %>%
  mutate(MntlEver_C = as.factor(MntlEver_C))

# suicide attempt (Suicide)
longitudinal_complete <- longitudinal_complete %>%
  mutate(SuicideF = as.factor(SuicideF)) %>%
  mutate(SuicideEver = ifelse(SuicideF %in% c(66, 77, 88, 99) | is.na(SuicideF), NA, SuicideF)) %>%
  mutate(
    SuicidePrior = ifelse(SuicidePrior %in% c(66, 77, 99) | is.na(SuicidePrior), NA, SuicidePrior),
    SuicideAttempt = case_when(
      is.na(SuicidePrior) ~ NA, 
      SuicidePrior == 88 ~ 1, # never attempted suicide
      SuicidePrior == 0 ~ 2, # yes, but not in the last year
      SuicidePrior == 1 ~ 3 # yes, in the last year
  )) %>%
  mutate(SuicideEver = as.factor(SuicideEver)) %>%
  mutate(SuicideAttempt = as.factor(SuicideAttempt))

# drinking category (DRINKCatF)
longitudinal_complete <- longitudinal_complete %>%
  mutate(DRINKCatF = as.factor(DRINKCatF)) %>%
  mutate(DRINKCatF_C = ifelse(DRINKCatF == 99 | is.na(DRINKCatF), NA, DRINKCatF)) %>%
  mutate(DRINKCatF_C = as.factor(DRINKCatF_C)) 

# problematic alcohol use
longitudinal_complete <- longitudinal_complete %>%
  mutate(PROBLEMUseF = as.factor(PROBLEMUseF)) %>%
  mutate(PROBLEMUseF_C = ifelse(PROBLEMUseF %in% c(77, 99) | is.na(PROBLEMUseF), NA, PROBLEMUseF)) %>% 
  mutate(PROBLEMUseF_C = as.factor(PROBLEMUseF_C))

# illicit drug use (DrugsF)
longitudinal_complete <- longitudinal_complete %>%
  mutate(DrugsF = as.factor(DrugsF)) %>%
  mutate(DrugsF_C = ifelse(DrugsF %in% c(66, 77, 88, 99) | is.na(DrugsF), NA, DrugsF)) %>%
  mutate(DrugsF_C = as.factor(DrugsF_C))

# depression (PHQ9TOTF)
longitudinal_complete <- longitudinal_complete %>%
  mutate(PHQ9TOTF_C= ifelse(PHQ9TOTF %in% c(666, 888, 999), NA, PHQ9TOTF))
longitudinal_complete <- longitudinal_complete %>%
  mutate(
    PHQPleasureF_C = if_else(PHQPleasureF %in% c(66, 81, 82, 99) | is.na(PHQPleasureF), NA, PHQPleasureF),
    PHQDownF_C = if_else(PHQDownF %in% c(66, 81, 82, 99) | is.na(PHQDownF), NA, PHQDownF),
    PHQ2Count = PHQPleasureF_C + PHQDownF_C,
    gate_PHQ = case_when(
      PHQ2Count == 0 ~ 1,
      PHQ2Count > 0 ~ 2,
      TRUE ~ NA
    ),
    gate_PHQ = as.factor(gate_PHQ))

# anxiety (GAD7TOTF)
longitudinal_complete <- longitudinal_complete %>%
  mutate(GAD7TOTF_C= ifelse(GAD7TOTF %in% c(888, 999), NA, GAD7TOTF))
longitudinal_complete <- longitudinal_complete %>%
  mutate(
    GADNervousF_C = if_else(GADNervousF %in% c(66, 81, 82, 99) | is.na(GADNervousF), NA, GADNervousF),
    GADCntrlWryF_C = if_else(GADCntrlWryF %in% c(66, 81, 82, 99) | is.na(GADCntrlWryF), NA, GADCntrlWryF),
    GAD2Count = GADNervousF_C + GADCntrlWryF_C,
    gate_GAD = case_when(
      GAD2Count == 0 ~ 1,
      GAD2Count > 0 ~ 2,
      TRUE ~ NA
    ),
    gate_GAD = as.factor(gate_GAD))

```


## Streamline

Streamline the dataset by selecting the "cleaned" version of each variable (ie, `Variable_C`).

```{r}
longitudinal_clean <- longitudinal_complete %>%
  select(Mod1Id, Mod2Id, FollowUpPeriod, AGENoPHIF_C, SexF_C, INJYEAR, AgeDeath_C, DEATHYEAR, DAYStoDEATHF_C, DeathCause1F, DeathCause2F, DeathECodeF, DeathAssault, DeathSuicide, DeathOverdose, EduYears10_C, EDUCATION_C, EarnF_C, HHIncomeF_C, Race1, RaceCat, EthnicityF_C, race_ethnicity, race_ethnicity_C, CountryBirthF_C, RURALF, EMPLOYMENTF_C, EMPLOY1, Cause, CauseCat, Aetiology, TBIInjuryF_C, TBIRepInjuryF_C, GCSCatSed, SCI_C, CC_Hypertension_C, DRSF_C, FIMTOTF_C, FIMCOGF_C, GOSTotalF_C, SpEd_C, PsyHosp_C, MntlEver_C, SuicideEver, SuicideAttempt, DRINKCatF_C, PROBLEMUseF, DrugsF_C, gate_PHQ, PHQ9TOTF_C, gate_GAD, GAD7TOTF_C)

# for analyses looking at baseline characteristics restrict the sample to those with Form 1 data
baseline_clean <- longitudinal_clean %>%
  filter(FollowUpPeriod == 0)
```

## Codebook

Create a codebook for all variables using the package `labelled()`.
```{r}
# Label your variables
var_label(longitudinal_clean) <- list(
  Mod1Id = "Unique identifier for rows in Form 1",
  Mod2Id = "Unique identifier for rows in Form 2", 
  FollowUpPeriod = "Follow up period, ie, data collection wave following index TBI", 
  AGENoPHIF_C = "Age at injury (unknown values were imputed based on age in Form 1)",
  SexF_C = "Sex. If transgender, current sex is recorded", 
  INJYEAR = "Year of injury", 
  AgeDeath_C = "Age at death (if age is unknown an estimated age at death based on death year - injury year is computed)", 
  DEATHYEAR = "Year of death", 
  DAYStoDEATHF_C = "Days from birth to death", 
  DeathCause1F = "Primary cause of death, defined using ICD codes", 
  DeathCause2F = "Secondary cause of death, defined using ICD codes", 
  DeathECodeF = "External cause of death, defined using ICD codes", 
  DeathAssault = "Binary variable indicating death by assault",
  DeathSuicide = "Binary variable indicating death by suicide", 
  DeathOverdose = "Binary variable indicating that the primary or secondary cause of death was related to substance misuse", 
  EduYears10_C = "Highest grade of school completed", 
  EDUCATION_C = "Calculated educational attainment",
  EarnF_C = "Annual income", 
  HHIncomeF_C = "Annual household income", 
  RaceCat = "Racial category", 
  Race1 = "Race where Pacific Islander and Native American are collapsed into one category due to small sample sizes",
  EthnicityF_C = "Self-reported ethnicity. Are you of Hispanic, Latino or Spanish origin?", 
  CountryBirthF_C = "Country of birth", 
  RURALF = "Urbanicity", 
  EMPLOYMENTF_C = "Employment status", 
  EMPLOY1 = "Calculated employment status with smaller categories",
  CauseCat = "Cause of injury", 
  Aetiology = "Cause of injury dichotimised to violent or nonviolent", 
  TBIInjuryF_C = "Is ther any head or neck injury reported", 
  TBIRepInjuryF_C = "Have you ever had a period of time in which you experienced multiple repeated impact to your head?", 
  GCSCatSed = "Glasgow Coma Score severity score", 
  SCI_C = "Spinal cord injury", 
  CC_Hypertension_C = "Intracranial hypertension", 
  DRSF_C = "Disability Rating Scale score", 
  FIMTOTF_C = "FIM total score", 
  FIMCOGF_C = "FIM cognitive scale score", 
  GOSTotalF_C = "Glasgow Outcome Scale total score", 
  SpEd_C = "In school, were you ever classified as a special education student", 
  PsyHosp_C = "Have you ever been hospitalised for a psychiatric problem?", 
  MntlEver_C = "Have you ever recievied treatment for any mental health problem", 
  SuicideEver = "Have you attempted suicide (lifetime)?", 
  SuicideAttempt = "Have you attempted suicide, and if so did you attempt in the last year?",
  DRINKCatF_C = "Alcohol consumption level", 
  PROBLEMUseF = "Substance problem use", 
  DrugsF_C = "Did you use any illict or non-prescription drugs in the last year?", 
  gate_PHQ = "Did you endorse either the pleasure of down items on the PHQ?", 
  PHQ9TOTF_C = "Patient Health Questionnaire (depression measure) total score", 
  gate_GAD = "Did you endorse either the pleasure of down items on the GAD?", 
  GAD7TOTF_C = "Generalised Anxiety Disorder (anxiety measure) total score"
)

codebook <- longitudinal_clean %>%
  look_for(details = "none") 

datatable(codebook, 
  options = list(
    scrollX = TRUE,             # enables horizontal scroll
    autoWidth = TRUE,           # adjusts column widths
    pageLength = 10             # optional: rows per page
  ),
  class = 'display nowrap stripe'
)
```

## Inclusion and exclusion criteria 

Most variables of interest are available in the dataset from October 2012. In order to reduce structural missingness, the sample was restricted to those enrolled in TBIMS between 2013 and 2023, reflecting the period when all variables of interest were available in the TBIMS NDB.

```{r}
data_risk <- baseline_clean %>% # <7>
  filter(INJYEAR >= 2013 & INJYEAR <= 2023) %>% # <8>
  select(-Mod2Id, -DAYStoDEATHF_C, -HHIncomeF_C, -gate_PHQ, -PHQ9TOTF_C, -gate_GAD, -GAD7TOTF_C, -TBIInjuryF_C, -TBIRepInjuryF_C, -GOSTotalF_C) # <9>
```

7.  As this study is a cross-sectional study I have filtered for the baseline/Form 1 data only.
8.  Restrict the sample to those enrolled in TBIMS between 2013 and 2023.
9.  Variables that only appear in Form 2, and thus are not available at baseline, are removed.


Relative to their younger counterparts, older adults more often sustain TBI from falls and less often from vehicular  or sports-related incidents and violence. We chose to restrict the analysis to those aged under 55 due to distinct injury mechanism patterns. 
```{r}
data_risk <- data_risk %>%
  filter(AGENoPHIF_C < 55)
```

Rename variables for ease of interpretation in results, now that the cohort is finalised. 
```{r}
data_risk <- data_risk %>%
  select(Mod1Id, CauseCat, Aetiology, PsyHosp_C, MntlEver_C, SuicideEver, SuicideAttempt, DrugsF_C, DRINKCatF_C, AGENoPHIF_C, SexF_C, race_ethnicity_C, RURALF, EDUCATION_C, SpEd_C, EMPLOY1, CountryBirthF_C, GCSCatSed)

data_risk %>%
  distinct(Mod1Id) %>%
  nrow()

data_risk <- data_risk %>%
  dplyr::rename(
    Age = AGENoPHIF_C,
    Sex = SexF_C,
    Race = race_ethnicity_C,
    Urbanicity = RURALF,
    Drugs = DrugsF_C,
    Suicide_Lifetime = SuicideEver,
    Suicide_Recent = SuicideAttempt,
    Special_Education = SpEd_C,
    Mental_Treatment = MntlEver_C, 
    Psychiatric_Hospitalisation = PsyHosp_C,
    Education = EDUCATION_C,
    Employment = EMPLOY1,
    GCS = GCSCatSed,
    Cause = CauseCat,
    Aetiology = Aetiology,
    Alcohol_Use = DRINKCatF_C,
    BornUS = CountryBirthF_C
  )
```

```{r}

data_risk <- data_risk %>%
  mutate(Sex = factor(Sex, labels = c("Female", "Male")),
         Race = factor(Race, labels = c("White", "Black or AA", "Asian / NHPI", "Hispanic or Latino", "AIAN", "Non-Hispanic Other", "More than one race")),
         BornUS = factor(BornUS, labels = c("Born in the US", "Born outside the US")),
         Urbanicity = factor(Urbanicity, labels = c("Rural", "Urban", "Suburban")),
         Special_Education = factor(Special_Education, labels = c("No", "Yes")),
         Employment = factor(Employment, labels = c("Competitively employed", "Student", "Retired", "Other employment", "Unemployed")),
         Education = factor(Education, labels = c("No high school degree", "HS/GED", "Some college", "Associates degree", "Bachelors degree", "Graduate degree")),
         Cause = factor(Cause, labels = c("Vehicular", "Gunshot wounds", "Assault with blunt instrument", "Other violence", "Sports", "Falls", "Other")),
         Aetiology = factor(Aetiology, labels = c("Nonviolent", "Violent")),
         Mental_Treatment = factor(Mental_Treatment, labels = c("No", "Yes")),
         Psychiatric_Hospitalisation = factor(Psychiatric_Hospitalisation, labels = c("No", "Yes")),
         Alcohol_Use = factor(Alcohol_Use, labels = c("Abstaining", "Light", "Moderate", "Heavy")),
         Suicide_Lifetime = factor(Suicide_Lifetime, labels = c("No", "Yes")),
         Suicide_Recent = factor(Suicide_Recent, labels = c("No", "Yes but not in the last year", "Yes and within the last year")),
         Drugs = factor(Drugs, labels = c("No", "Yes")),
         GCS = factor(GCS, labels = c("Severe", "Moderate", "Mild", "Sedated", "Intubated"))
         )
```


## Missing data

### Visualise missingness
```{r, results='hide'}
data_risk %>%
  select(-Mod1Id) %>%
  gg_miss_var(show_pct = TRUE) + labs(title = "Proportion of missingness")

jpeg("output/naniar_missing.jpg", width = 800, height = 600)
data_risk %>%
  select(-Mod1Id) %>%
  gg_miss_var(show_pct = TRUE) + labs(title = "Proportion of missingness")
dev.off()
```

```{r, fig.height= 15}
par(mar = c(10, 6, 4, 2))
aggr(data_risk, 
     numbers = TRUE, 
     prop = FALSE, 
     sortVar = TRUE, 
     combined = TRUE, 
     ylab = "Missing Data Pattern",
     cex.lab = 1.2,
     cex.axis = 0.4)

jpeg("output/patterns_missing_VIM.jpg", width = 800, height = 1500)
aggr(data_risk, 
     numbers = TRUE, 
     prop = FALSE, 
     sortVar = TRUE, 
     combined = TRUE, 
     ylab = "Missing Data Pattern",
     cex.lab = 1.2,
     cex.axis = 0.4)
dev.off()
```

```{r}
par(mar = c(10, 6, 4, 2))
aggr(data_risk %>%
       select(Alcohol_Use, Drugs, Psychiatric_Hospitalisation, Mental_Treatment, Suicide_Lifetime, Suicide_Recent),
     numbers = TRUE, 
     prop = FALSE, 
     sortVar = TRUE, 
     combined = FALSE, 
     cex.lab = 1.2,
     cex.axis = 0.4)
```

#### Assess the patterns of missing data

Does missingness for race/ethnicity vary by observed variables? 
```{r}
data_sub <- data_risk %>%
  mutate(
    # categorise age
    AgeCat = case_when(
    Age < 21 ~ "16-20y",
    Age >= 21 & Age < 26 ~ "21-25y",
    Age >= 26 & Age < 31 ~ "26-30y",
    Age >= 31 & Age < 41 ~ "31-40y",
    Age >= 41 & Age < 55 ~ "41-54y"
  ),
   
  # is alcohol use missing?
    Race_Missing = ifelse(is.na(Race), "Yes", "No")) %>%
  select(AgeCat, Sex, BornUS, Urbanicity, Aetiology, Special_Education, Employment, Education, Race_Missing, Mental_Treatment, Drugs)


data_sub %>%
  tbl_summary(by = Race_Missing, include = c(AgeCat, Sex, BornUS, Urbanicity, Special_Education, Employment, Education)) %>%
  modify_spanning_header(all_stat_cols() ~ "**Missing Race/Ethnicity**") %>%
  add_p(
    # perform  chi-square with simulation to avoid computational constraints
    test = all_categorical() ~ "chisq.test",
test.args = all_categorical() ~ list(simulate.p.value = TRUE))
```

```{r}
tf <- "output/gtsummary-race-missing.docx"

data_sub %>%
  tbl_summary(by = Race_Missing, include = c(AgeCat, Sex, BornUS, Urbanicity, Special_Education, Employment, Education)) %>%
  modify_spanning_header(all_stat_cols() ~ "**Missing Race/Ethnicity**") %>%
  add_p(
    # perform  chi-square with simulation to avoid computational constraints
    test = all_categorical() ~ "chisq.test",
test.args = all_categorical() ~ list(simulate.p.value = TRUE)) %>%
  as_gt() %>%
  gt::gtsave(filename = tf)
```

```{r}
age.race <- data_risk[, c("Race", "Age")]
marginplot(age.race)

jpeg("output/age_race_marginplot.jpg", width = 800, height = 600)
marginplot(age.race)
dev.off()
```

```{r}
race_select <- data.frame(data_risk[,c("Race", "Age", "BornUS", "Education", "Employment")])

# copy variable with missing data and rename 
race_select$Race_MAR <- race_select$Race

explanatory = c("Age",  "Age", "BornUS", "Education", "Employment")
dependent = "Race_MAR"
race_select %>% 
  missing_compare(dependent, explanatory) %>% 
    knitr::kable(row.names=FALSE, align = c("l", "l", "r", "r", "r"), 
        caption = "Mean comparisons between values of responders (Not missing) and 
        non-responders (Missing) on the race/ethnicity variable.") 
```


Does missingness for alcohol use vary by observed variables? 

```{r}
data_sub <- data_risk %>%
  mutate(
    # categorise age
    AgeCat = case_when(
    Age < 21 ~ "16-20y",
    Age >= 21 & Age < 26 ~ "21-25y",
    Age >= 26 & Age < 31 ~ "26-30y",
    Age >= 31 & Age < 41 ~ "31-40y",
    Age >= 41 & Age < 55 ~ "41-54y"
  ),
   
  # is alcohol use missing?
    Alcohol_Missing = ifelse(is.na(Alcohol_Use), "Yes", "No")) %>%
  select(AgeCat, Sex, Race, BornUS, Urbanicity, Aetiology, Special_Education, Employment, Education, Alcohol_Missing, Mental_Treatment, Drugs)


data_sub %>%
  tbl_summary(by = Alcohol_Missing, include = c(AgeCat, Sex, Race, BornUS, Urbanicity, Special_Education, Employment, Education, Mental_Treatment, Drugs)) %>%
  modify_spanning_header(all_stat_cols() ~ "**Missing Alcohol Consumption**") %>%
  add_p(
    # perform  chi-square with simulation to avoid computational constraints
    test = all_categorical() ~ "chisq.test",
test.args = all_categorical() ~ list(simulate.p.value = TRUE))
```

```{r}
tf2 <- "output/gtsummary-alcohol-missing.docx"

data_sub %>%
  tbl_summary(by = Alcohol_Missing, include = c(AgeCat, Sex, Race, BornUS, Urbanicity, Special_Education, Employment, Education, Mental_Treatment, Drugs)) %>%
  modify_spanning_header(all_stat_cols() ~ "**Missing Alcohol Consumption**") %>%
  add_p(
    # perform  chi-square with simulation to avoid computational constraints
    test = all_categorical() ~ "chisq.test",
test.args = all_categorical() ~ list(simulate.p.value = TRUE)) %>%
  as_gt() %>%
  gt::gtsave(filename = tf2)
```

```{r}
age.alc <- data_risk[, c("Alcohol_Use", "Age")]
marginplot(age.alc)

jpeg("output/age_alcohol_marginplot.jpg", width = 800, height = 600)
marginplot(age.alc)
dev.off()
```

```{r}
alc_select <- data.frame(data_risk[,c("Alcohol_Use", "Age", "Drugs", "Mental_Treatment", "Psychiatric_Hospitalisation", "Suicide_Lifetime", "Suicide_Recent")])

# copy variable with missing data and rename 
alc_select$Alcohol_MAR <- alc_select$Alcohol_Use

explanatory = c("Age", "Drugs", "Mental_Treatment", "Psychiatric_Hospitalisation", "Suicide_Lifetime", "Suicide_Recent")
dependent = "Alcohol_MAR"
alc_select %>% 
  missing_compare(dependent, explanatory) %>% 
    knitr::kable(row.names=FALSE, align = c("l", "l", "r", "r", "r"), 
        caption = "Mean comparisons between values of responders (Not missing) and 
        non-responders (Missing) on the alcohol use variable.")
```

```{r}
drug.alc <- as.data.frame(data_risk[, c("Alcohol_Use", "Drugs")])
spineMiss(drug.alc)

jpeg("output/drug_alcohol_spineplot.jpg", width = 800, height = 600)
spineMiss(drug.alc)
dev.off()
```

### Multiple imputation

#### Fitting the imputed model

Include all variables that will be used in the analysis in the same form they will be used in the model. 
```{r}
data_risk <- data_risk %>%
  select(-Aetiology) # <5> I don't include `Aetiology` in the imputation model, as `Aetiology` is a collapsed version of `Cause` and including both causes perfect/near-perfect multicollinearity

imp.mids <- mice(data_risk,
                   seed  = 3, # ensure reproducibility
                   m     = 2, # when you are ready to fit your final imputation model, use 20 imputations
                   print = F)
```

```{r}
summary(imp.mids)
```


Return the imputed dataset using `complete()`.
```{r}
imp.data <- complete(imp.mids)  # return the first imputed dataset
```

Does the imputed dataset have the same dimensions as the original data?
```{r}
dim(data_risk)
dim(imp.data)
```

Return all imputed datasets in "long" format (with stacking) including the original data
```{r}
imp.data <- complete(imp.mids, "long", include = TRUE)
class(imp.data)
dim(imp.data)
```

```{r, results='hide'}
names(imp.data) #<4>

range(imp.data$.imp)
range(imp.data$.id)
```

4. A new variable called ".imp" indexes the imputations, where .imp = 0 corresponds to the original data and .imp = 1 to m corresponds to th imputed data. The new variable ".id" indexes the observations 

Dichotimse for the outcome (violent vs nonviolent TBI)
```{r}
imp.data <- imp.data %>%
  mutate(
    Aetiology = case_when(
    Cause == "Vehicular" | Cause == "Sports" | Cause == "Falls" | Cause == "Other" ~ 0,
    Cause == "Gunshot wounds" | Cause == "Assault with blunt instrument" | Cause == "Other violence" ~ 1,
  ),
  Aetiology = factor(Aetiology, labels = c("Nonviolent", "Violent")))

imp.risk <- as.mids(imp.data, where = NULL, .imp = ".imp", .id = ".id")
```


#### Examine the imputed values

Create a table comparing the distribution of the variable between the complete data and the imputed data.

```{r}
imp.data <- imp.data %>% 
  mutate(imputed = .imp > 0,
         imputed = factor(imputed,
                          levels = c(FALSE, TRUE),
                          labels = c("Observed", "Imputed")))

prop_observed_imputed <- function(data, vars, imp_var) {

  data %>%
    select(all_of(c(vars, imp_var))) %>%
    tbl_summary(
      by = !!sym(imp_var),
      statistic = all_categorical() ~ "{p}%",
      percent = "column",
      missing = "no"
    ) %>%
    add_p() %>%
    bold_labels()
}

prop_observed_imputed(
  data   = imp.data,
  vars   = c("Sex", "Race", "BornUS", "Urbanicity", "Employment", "Education",
             "Special_Education", "GCS", "Cause", "Aetiology",
             "Psychiatric_Hospitalisation", "Mental_Treatment",
             "Suicide_Lifetime", "Suicide_Recent", "Alcohol_Use", "Drugs"),
  imp_var = "imputed"
)

```


#### Complete case analysis

How many rows would we lose if we required complete cases for these predictors?
```{r}
data_complete <- data_risk

data_miss <- data_complete

n_total <- nrow(data_miss)
n_cc    <- data_miss %>%
  complete.cases() %>%
  sum()

cbind(
  total_rows    = n_total,
  complete_cases= n_cc,
  lost_rows     = n_total - n_cc,
  lost_pct      = round((n_total - n_cc) / n_total * 100, 1)
)
```

# Descriptive statistics

After MI, compute descriptive statistics within each complete dataset and use Rubin’s rules to **pool** statistics and compute their standard errors.

Use the `mi.mean.se.sd()` function (found in `Functions_rmph.R`) to calculate descriptive for the continuous variable `AGENoPHI`. The functions within `Functions_rmph.R` operate on a `mids` object. 

```{r}
# use the original 'mids' object imp.risk
mi.mean.se.sd(imp.risk, "Age")

# frequencies and proportions
rbind(mi.n.p(imp.risk, "Aetiology"),
      mi.n.p(imp.risk, "Sex"),
      mi.n.p(imp.risk, "Race"),
      mi.n.p(imp.risk, "Urbanicity"),
      mi.n.p(imp.risk, "BornUS"),
      mi.n.p(imp.risk, "Education"),
      mi.n.p(imp.risk, "Special_Education"),
      mi.n.p(imp.risk, "Employment"),
      mi.n.p(imp.risk, "GCS"),
      mi.n.p(imp.risk, "Cause"),
      mi.n.p(imp.risk, "Mental_Treatment"),
      mi.n.p(imp.risk, "Psychiatric_Hospitalisation"),
      mi.n.p(imp.risk, "Suicide_Lifetime"),
      mi.n.p(imp.risk, "Alcohol_Use"),
      mi.n.p(imp.risk, "Drugs")
      )
```

By TBI aetiology (`Aetiology`)
```{r}
# use the original 'mids' object imp.risk
mi.mean.se.sd.by(imp.risk, "Age", BY = "Aetiology")

# frequencies and proportions
rbind(mi.n.p.by(imp.risk, "Sex", BY = "Aetiology"),
      mi.n.p.by(imp.risk, "Race", BY = "Aetiology"),
      mi.n.p.by(imp.risk, "Urbanicity", BY = "Aetiology"),
      mi.n.p.by(imp.risk, "BornUS", BY = "Aetiology"),
      mi.n.p.by(imp.risk, "Education", BY = "Aetiology"),
      mi.n.p.by(imp.risk, "Special_Education", BY = "Aetiology"),
      mi.n.p.by(imp.risk, "Employment", BY = "Aetiology"),
      mi.n.p.by(imp.risk, "GCS", BY = "Aetiology"),
      mi.n.p.by(imp.risk, "Mental_Treatment", BY = "Aetiology"),
      mi.n.p.by(imp.risk, "Psychiatric_Hospitalisation", BY = "Aetiology"),
      mi.n.p.by(imp.risk, "Suicide_Lifetime", BY = "Aetiology"),
      mi.n.p.by(imp.risk, "Alcohol_Use", BY = "Aetiology"),
      mi.n.p.by(imp.risk, "Drugs", BY = "Aetiology")
      )
```

# Logistic regression

## Linearity assumptions

The sociodemographic factor `Age` is a continuous variable, and therefore the linearity assumption must be explored. Lets assess this before building the logistic regression models using restricted cubic splines. 
```{r}
# before fitting models, remove missing values
data_risk <- data_risk %>%
  mutate(Aetiology = case_when(
    Cause == "Vehicular" | Cause == "Sports" | Cause == "Falls" | Cause == "Other" ~ 0,
    Cause == "Gunshot wounds" | Cause == "Assault with blunt instrument" | Cause == "Other violence" ~ 1,
    TRUE ~ NA
  )) 

table(data_risk$Aetiology)

baseline_complete <- data_risk[complete.cases(data_risk$Age, 
                                                    data_risk$Aetiology), ]

# Now fit models on complete data
agebaseline3_spline <- rcs(baseline_complete$Age, 3)
agebaseline4_spline <- rcs(baseline_complete$Age, 4)
agebaseline5_spline <- rcs(baseline_complete$Age, 5)

# fit logistic regression models
logit_baseline3 <- glm(Aetiology ~ agebaseline3_spline, 
                       data = baseline_complete, 
                       family = binomial(link = "logit"))

logit_baseline4 <- glm(Aetiology ~ agebaseline4_spline, 
                       data = baseline_complete,
                       family = binomial(link = "logit"))

logit_baseline5 <- glm(Aetiology ~ agebaseline5_spline, 
                       data = baseline_complete, 
                       family = binomial(link = "logit"))

# get linear predictors
lp_baseline3 <- predict(logit_baseline3, type = "link")
lp_baseline4 <- predict(logit_baseline4, type = "link")
lp_baseline5 <- predict(logit_baseline5, type = "link")

# create data frames for plotting
data_splines <- data.frame(
  age_at_entry = baseline_complete$Age,
  lp_baseline3 = lp_baseline3,
  lp_baseline4 = lp_baseline4,
  lp_baseline5 = lp_baseline5
)

# reshape to long format
data_splines_m <- melt(data_splines, id.vars = 'age_at_entry')

# plot
ggplot(data_splines_m, 
                       aes(x=age_at_entry, y=value, colour=variable)) + 
  geom_line() + 
   scale_colour_manual(labels = c("3 knots", "4 knots", "5 knots"), values = c("darkseagreen", 'cadetblue3', "deeppink3")) + 
  theme_brand() +
  labs(x = "Age", 
       y = "Linear Prediction (Log-Odds)", 
       color = "") + 
  theme(legend.position = c(0.8, 0.2))

jpeg("output/age_rcs.jpg", width = 800, height = 600)
ggplot(data_splines_m, 
                       aes(x=age_at_entry, y=value, colour=variable)) + 
  geom_line() + 
   scale_colour_manual(labels = c("3 knots", "4 knots", "5 knots"), values = c("darkseagreen", 'cadetblue3', "deeppink3")) + 
  theme_brand() +
  labs(x = "Age", 
       y = "Linear Prediction (Log-Odds)", 
       color = "") + 
  theme(legend.position = c(0.8, 0.2))
dev.off()
```

```{r}
model_comparison <- data.frame(
  Model = c("logit_baseline3", "logit_baseline4", "logit_baseline5"),
  Knots = c(3, 4, 5),  # or whatever your knot values are
  AIC = c(
    AIC(logit_baseline3),
    AIC(logit_baseline4),
    AIC(logit_baseline5)
  ),
  BIC = c(
    BIC(logit_baseline3),
    BIC(logit_baseline4),
    BIC(logit_baseline5)
  )
)

model_comparison %>%
  gt() %>%
  tab_header(
    title = "Model Comparison",
    subtitle = "AIC and BIC Values"
  ) %>%
  fmt_number(
    columns = c(AIC, BIC),
    decimals = 2
  ) %>%
  # Highlight minimum AIC in green
  tab_style(
    style = cell_fill(color = "lightblue3"),
    locations = cells_body(
      columns = AIC,
      rows = AIC == min(AIC)
    )
  ) %>%
  # Highlight minimum BIC in green
  tab_style(
    style = cell_fill(color = "lightblue3"),
    locations = cells_body(
      columns = BIC,
      rows = BIC == min(BIC)
    )
  )
```

Age at injury should be modelled using restricted cubic splines with 3 knots.

## Minimally adjusted models
```{r}
imp.psych_rcs <- with(imp.risk, 
                      glm(Aetiology ~ Psychiatric_Hospitalisation + rcs(Age, 3) + Sex, family = binomial))
```

```{r}
tbl_regression(imp.psych_rcs, 
               exponentiate = TRUE) %>% # Shows ORs instead of log-odds
  bold_p() %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  as_gt() %>%  # Convert to gt for further customization
  gt::tab_header(
    title = "Logistic Regression: TBI Aetiology by Psychiatric Hospitalisation"
  ) 

```

```{r}
imp.mntl_rcs <- with(imp.risk, 
                      glm(Aetiology ~ Mental_Treatment + rcs(Age, 3) + Sex, family = binomial))
```

```{r}
tbl_regression(imp.mntl_rcs, 
               exponentiate = TRUE) %>%
  bold_p() %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  as_gt() %>%  # Convert to gt for further customization
  gt::tab_header(
    title = "Logistic Regression: TBI Aetiology by Mental Health Treatment"
  )
```

```{r}
imp.suicide_rcs <- with(imp.risk, 
                      glm(Aetiology ~ Suicide_Lifetime + rcs(Age, 3) + Sex, family = binomial))
```

```{r}
tbl_regression(imp.suicide_rcs, 
               exponentiate = TRUE) %>%
  bold_p() %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  as_gt() %>%  # Convert to gt for further customization
  gt::tab_header(
    title = "Logistic Regression: TBI Aetiology by Suicide Attempt (Lifetime)"
  )
```

```{r}
imp.drink_rcs <- with(imp.risk, 
                      glm(Aetiology ~ Alcohol_Use + rcs(Age, 3) + Sex, family = binomial))
```

```{r}
tbl_regression(imp.drink_rcs, 
               exponentiate = TRUE) %>%
  bold_p() %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  as_gt() %>%  # Convert to gt for further customization
  gt::tab_header(
    title = "Logistic Regression: TBI Aetiology by Alcohol Use"
  )
```

```{r}
imp.drug_rcs <- with(imp.risk, 
                      glm(Aetiology ~ Drugs + rcs(Age, 3) + Sex, family = binomial))

```

```{r}
tbl_regression(imp.drug_rcs, 
               exponentiate = TRUE) %>%
  bold_p() %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  as_gt() %>%  # Convert to gt for further customization
  gt::tab_header(
    title = "Logistic Regression: TBI Aetiology by Illicit Drug Use"
  )
```


## Fully adjusted models 

Adjust for all sociodemographic risk factors (age, sex, race, urbanicity, educational attainment, learning difficulties/special education, and employment status) in the logistic regression models.

```{r}
imp.psych.full <- with(imp.risk, 
                      glm(Aetiology ~ Psychiatric_Hospitalisation + rcs(Age, 3) + Sex
                          + Race + Urbanicity + Education + Special_Education + Employment, family = binomial))
```

```{r}
tbl_regression(imp.psych.full, 
               exponentiate = TRUE) %>%
  bold_p() %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  as_gt() %>%  # Convert to gt for further customization
  gt::tab_header(
    title = "Logistic Regression: TBI Aetiology by Psychiatric Hospitalisation (Fully Adjusted Model)"
  )
```


```{r}
imp.mntl.full <- with(imp.risk, 
                      glm(Aetiology ~ Mental_Treatment + rcs(Age, 3) + Sex
                           + Race + Urbanicity + Education + Special_Education + Employment, family = binomial))
```

```{r}
tbl_regression(imp.mntl.full, 
               exponentiate = TRUE) %>%
  bold_p() %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  as_gt() %>%  # Convert to gt for further customization
  gt::tab_header(
    title = "Logistic Regression: TBI Aetiology by Mental Health Treatment (Fully Adjusted Model)"
  )
```

```{r}
imp.suicide.full <- with(imp.risk, 
                      glm(Aetiology ~ Suicide_Lifetime + rcs(Age, 3) + Sex
                           + Race + Urbanicity + Education + Special_Education + Employment, family = binomial))
```

```{r}
tbl_regression(imp.suicide.full, 
               exponentiate = TRUE) %>%
  bold_p() %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  as_gt() %>%  # Convert to gt for further customization
  gt::tab_header(
    title = "Logistic Regression: TBI Aetiology by Suicide Attempt (Fully Adjusted Model)"
  )
```

```{r}
imp.drink.full <- with(imp.risk, 
                      glm(Aetiology ~ Alcohol_Use + rcs(Age, 3) + Sex
                           + Race + Urbanicity + Education + Special_Education + Employment, family = binomial))
```

```{r}
tbl_regression(imp.drink.full, 
               exponentiate = TRUE) %>%
  bold_p() %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  as_gt() %>%  # Convert to gt for further customization
  gt::tab_header(
    title = "Logistic Regression: TBI Aetiology by Alcohol Use (Fully Adjusted Model)"
  )
```

```{r}
imp.drug.full <- with(imp.risk, 
                      glm(Aetiology ~ Drugs + rcs(Age, 3) + Sex
                           + Race + Urbanicity + Education + Special_Education + Employment, family = binomial))
```

```{r}
tbl_regression(imp.drug.full, 
               exponentiate = TRUE) %>%
  bold_p() %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  as_gt() %>%  # Convert to gt for further customization
  gt::tab_header(
    title = "Logistic Regression: TBI Aetiology by Illicit Drug Use (Fully Adjusted Model)"
  )
```

## All psychiatric predictors

### Correlation matrix

First, create a heat map of all psychiatric risk factors. Correlation of dark green indicates a 1.0 correlation. 
```{r}
coul <- colorRampPalette(brewer.pal(8, "PiYG"))(25)

completed <- complete(imp.risk, action = "all")

completed_mat <- lapply(completed, function(df) {
  as.matrix(sapply(df, as.numeric))
})

pooled_data <- Reduce("+", completed_mat) / length(completed_mat)

pooled_psych <- pooled_data[, c("Psychiatric_Hospitalisation", "Mental_Treatment", "Suicide_Lifetime", "Drugs", "Alcohol_Use")]

cor <- cor(pooled_psych, use = "pairwise.complete.obs", method = "pearson")

heatmap(cor, col = coul, symm = TRUE)

jpeg("output/heatmap.jpg", width = 800, height = 600)
heatmap(cor, col = coul, symm = TRUE)
dev.off()
```

The Phi coefficient is a measure of association between binary variables. A pearson correlation coefficient estimates for binary variables will return the phi coefficient.
```{r}
res <- round(cor, 2)
datatable(res, 
  options = list(
    scrollX = TRUE,             # enables horizontal scroll
    autoWidth = TRUE,           # adjusts column widths
    pageLength = 10             # optional: rows per page
  ),
  class = 'display nowrap stripe'
)
```


### Regression with all psychiatric predictors

Include all psychiatric predictors in the model, adjusting for sociodemographics covariates

```{r}
imp.psych.all <- with(imp.risk, 
                      glm(Aetiology ~ Psychiatric_Hospitalisation + Mental_Treatment +
                            Suicide_Lifetime + Alcohol_Use + Drugs + 
                            rcs(Age, 3) + Sex + 
                            Race + Urbanicity + Education + Special_Education + Employment, 
                          family = binomial))
```

```{r}
tbl_regression(imp.psych.all, 
               exponentiate = TRUE) %>%
  bold_p() %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  as_gt() %>%  # Convert to gt for further customization
  gt::tab_header(
    title = "Logistic Regression With All Psychiatric Predictors (Fully Adjusted Model)"
  )
```

## Assessing for multicollinearity 

An important step in the logistic regression analysis is to assess the assumption of no strong multicollinearity. To do so, we used the `car::viff()` function to calculate the variance inflation factor (VIF), which was pooled across all imputed datasets. 

For the `car::viff()` output, the generalized VIF is found in the GVIF column, and the GVIF^(1/(2*Df)) column is the adjusted generalized standard error inflation factor (aGSIF). It is recommended that the aGSIF value to evaluate collinearity, since it allows predictors with different Df (different number of terms in the model) to be comparable [(Nahhas, 2024)](https://bookdown.org/rwnahhas/RMPH/mlr-collinearity.html#mlr-collinearity).

For aGSIFs, our rule of thumb for what is large is that values **above 1.6 may be of concern** and values above 2.2 or 3.2 are indicative of a more serious problem [(Nahhas, 2024)](https://bookdown.org/rwnahhas/RMPH/mlr-collinearity.html#mlr-collinearity).

### Minimally adjusted models

Psychiatric Hospitalisation
```{r}
vif_results_psy <- lapply(1:imp.risk$m, function(i) {
  complete_data <- complete(imp.risk, i)
  imp.psych_rcs <- glm(Aetiology ~ Psychiatric_Hospitalisation + rcs(Age, 3) + Sex, 
               data = complete_data, 
               family = binomial)
  car::vif(imp.psych_rcs)
})

# average VIF across imputations
vif_averaged1 <- Reduce("+", vif_results_psy) / length(vif_results_psy)
print("Average VIF across imputations:")
print(vif_averaged1)
```

Mental Health Treatment
```{r}
vif_results_mnt <- lapply(1:imp.risk$m, function(i) {
  complete_data <- complete(imp.risk, i)
  imp.mntl_rcs <- glm(Aetiology ~ Mental_Treatment + rcs(Age, 3) + Sex, 
               data = complete_data, 
               family = binomial)
  car::vif(imp.mntl_rcs)
})

# average VIF across imputations
vif_averaged2 <- Reduce("+", vif_results_mnt) / length(vif_results_mnt)
print("Average VIF across imputations:")
print(vif_averaged2)
```

Suicide Attempt
```{r}
vif_results_suicide <- lapply(1:imp.risk$m, function(i) {
  complete_data <- complete(imp.risk, i)
  imp.suicide_rcs <- glm(Aetiology ~ Suicide_Lifetime + rcs(Age, 3) + Sex, 
               data = complete_data, 
               family = binomial)
  car::vif(imp.suicide_rcs)
})

# average VIF across imputations
vif_averaged3 <- Reduce("+", vif_results_suicide) / length(vif_results_suicide)
print("Average VIF across imputations:")
print(vif_averaged3)
```

Alcohol Use
```{r}
vif_results_drink <- lapply(1:imp.risk$m, function(i) {
  complete_data <- complete(imp.risk, i)
  imp.drink_rcs <- glm(Aetiology ~ Alcohol_Use + rcs(Age, 3) + Sex, 
               data = complete_data, 
               family = binomial)
  car::vif(imp.drink_rcs)
})

# average VIF across imputations
vif_averaged4 <- Reduce("+", vif_results_drink) / length(vif_results_drink)
print("Average VIF across imputations:")
print(vif_averaged4)
```

Illicit Drug Use
```{r}
vif_results_drugs <- lapply(1:imp.risk$m, function(i) {
  complete_data <- complete(imp.risk, i)
  imp.drug_rcs <- glm(Aetiology ~ Drugs + rcs(Age, 3) + Sex, 
               data = complete_data, 
               family = binomial)
  car::vif(imp.drug_rcs)
})

# average VIF across imputations
vif_averaged5 <- Reduce("+", vif_results_drugs) / length(vif_results_drugs)
print("Average VIF across imputations:")
print(vif_averaged5)
```

### Fully adjusted models

Psychiatric Hospitilisation
```{r}
vif_results_psy_full <- lapply(1:imp.risk$m, function(i) {
  complete_data <- complete(imp.risk, i)
   imp.psych.full <- glm(Aetiology ~ Psychiatric_Hospitalisation + rcs(Age, 3) + Sex
                          + Race + Urbanicity + Education + Special_Education + Employment,
               data = complete_data, 
               family = binomial)
  car::vif(imp.psych.full)
})

# average VIF across imputations
vif_averaged6 <- Reduce("+", vif_results_psy_full) / length(vif_results_psy_full)
print("Average VIF across imputations:")
print(vif_averaged6)
```

Mental Health Treatment 
```{r}
vif_results_mntl_full <- lapply(1:imp.risk$m, function(i) {
  complete_data <- complete(imp.risk, i)
   imp.mntl.full <- glm(Aetiology ~ Mental_Treatment + rcs(Age, 3) + Sex
                          + Race + Urbanicity + Education + Special_Education + Employment,
               data = complete_data, 
               family = binomial)
  car::vif(imp.mntl.full)
})

# average VIF across imputations
vif_averaged7 <- Reduce("+", vif_results_mntl_full) / length(vif_results_mntl_full)
print("Average VIF across imputations:")
print(vif_averaged7)
```

Suicide Attempt
```{r}
vif_results_suicide_full <- lapply(1:imp.risk$m, function(i) {
  complete_data <- complete(imp.risk, i)
   imp.suicide.full <- glm(Aetiology ~ Suicide_Lifetime + rcs(Age, 3) + Sex
                          + Race + Urbanicity + Education + Special_Education + Employment,
               data = complete_data, 
               family = binomial)
  car::vif(imp.suicide.full)
})

# average VIF across imputations
vif_averaged8 <- Reduce("+", vif_results_suicide_full) / length(vif_results_suicide_full)
print("Average VIF across imputations:")
print(vif_averaged8)
```

Alcohol Use
```{r}
vif_results_drink_full <- lapply(1:imp.risk$m, function(i) {
  complete_data <- complete(imp.risk, i)
   imp.drink.full <- glm(Aetiology ~ Alcohol_Use + rcs(Age, 3) + Sex
                          + Race + Urbanicity + Education + Special_Education + Employment,
               data = complete_data, 
               family = binomial)
  car::vif(imp.drink.full)
})

# average VIF across imputations
vif_averaged9 <- Reduce("+", vif_results_drink_full) / length(vif_results_drink_full)
print("Average VIF across imputations:")
print(vif_averaged9)
```

Illicit Drug Use
```{r}
vif_results_drugs_full <- lapply(1:imp.risk$m, function(i) {
  complete_data <- complete(imp.risk, i)
   imp.drugs.full <- glm(Aetiology ~ Drugs + rcs(Age, 3) + Sex
                          + Race + Urbanicity + Education + Special_Education + Employment,
               data = complete_data, 
               family = binomial)
  car::vif(imp.drugs.full)
})

# average VIF across imputations
vif_averaged10 <- Reduce("+", vif_results_drugs_full) / length(vif_results_drugs_full)
print("Average VIF across imputations:")
print(vif_averaged10)
```

All predictors
```{r}
vif_results_all <- lapply(1:imp.risk$m, function(i) {
  complete_data <- complete(imp.risk, i)
   imp.psych.all <- glm(Aetiology ~ Psychiatric_Hospitalisation + Mental_Treatment + Suicide_Lifetime + Alcohol_Use + Drugs
                          + rcs(Age, 3) + Sex
                          + Race + Urbanicity + Education + Special_Education + Employment,
               data = complete_data, 
               family = binomial)
  car::vif(imp.psych.all)
})

# average VIF across imputations
vif_averaged11 <- Reduce("+", vif_results_all) / length(vif_results_all)
print("Average VIF across imputations:")
print(vif_averaged11)
```

When including all psychiatric predictors the aGSIF increase, however still the values are still above 1.6 and therefore, shouldn't raise any cause for concern. 

# Sensitivity analysis

Are similar results obtained when the analysis is restricted to the complete cases only?

```{r}
data_complete <- data_risk

data_complete <- data_complete[complete.cases(data_complete[ , c("Aetiology", "Age", "Sex", "Race", "Urbanicity", "Drugs", "Suicide_Lifetime", "Suicide_Recent", "Special_Education", "Mental_Treatment", "Psychiatric_Hospitalisation", "Education", "Employment", "GCS", "Cause" , "Alcohol_Use", "BornUS")]), ]

data_complete %>%
  distinct(Mod1Id) %>%
  nrow()
```

## Descriptive statistics 

```{r}
data_complete %>%
  tbl_summary(by = Aetiology, include = c(Age, Sex, Race, BornUS, Urbanicity, Employment, Education, Special_Education, Psychiatric_Hospitalisation, Mental_Treatment, Suicide_Lifetime, Alcohol_Use, Drugs)) %>%
  add_p(
    # perform  chi-square with simulation to avoid computational constraints
    test = all_categorical() ~ "chisq.test",
test.args = all_categorical() ~ list(simulate.p.value = TRUE)) %>%
  add_overall()
```

Aetiology by psychiatric hospitalisation 
```{r}
data_complete %>%
  tbl_summary(by = Psychiatric_Hospitalisation, include = Aetiology) %>%
  add_p(
    # perform  chi-square with simulation to avoid computational constraints
    test = all_categorical() ~ "chisq.test",
test.args = all_categorical() ~ list(simulate.p.value = TRUE)) %>%
  add_overall()
```

Aetiology by mental health treatment 
```{r}
data_complete %>%
  tbl_summary(by = Mental_Treatment, include = Aetiology) %>%
  add_p(
    # perform  chi-square with simulation to avoid computational constraints
    test = all_categorical() ~ "chisq.test",
test.args = all_categorical() ~ list(simulate.p.value = TRUE)) %>%
  add_overall()
```

Aetiology by suicide attempt
```{r}
data_complete %>%
  tbl_summary(by = Suicide_Lifetime, include = Aetiology) %>%
  add_p(
    # perform  chi-square with simulation to avoid computational constraints
    test = all_categorical() ~ "chisq.test",
test.args = all_categorical() ~ list(simulate.p.value = TRUE)) %>%
  add_overall()
```

Aetiology by alcohol use
```{r}
data_complete %>%
  tbl_summary(by = Alcohol_Use, include = Aetiology) %>%
  add_p(
    # perform  chi-square with simulation to avoid computational constraints
    test = all_categorical() ~ "chisq.test",
test.args = all_categorical() ~ list(simulate.p.value = TRUE)) %>%
  add_overall()
```

Aetiology by illicit drug use
```{r}
data_complete %>%
  tbl_summary(by = Drugs, include = Aetiology) %>%
  add_p(
    # perform  chi-square with simulation to avoid computational constraints
    test = all_categorical() ~ "chisq.test",
test.args = all_categorical() ~ list(simulate.p.value = TRUE)) %>%
  add_overall()
```


## Minimally adjusted models
```{r}
cc.psych_rcs <- glm(Aetiology ~ Psychiatric_Hospitalisation + rcs(Age, 3) + Sex, family = binomial, data = data_complete)
```

```{r}
tbl_regression(cc.psych_rcs, 
               exponentiate = TRUE) %>% # Shows ORs instead of log-odds
  bold_p() %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  as_gt() %>%  # Convert to gt for further customization
  gt::tab_header(
    title = "Logistic Regression: TBI Aetiology by Psychiatric Hospitalisation"
  ) 

```

```{r}
cc.mntl_rcs <- glm(Aetiology ~ Mental_Treatment + rcs(Age, 3) + Sex, family = binomial, data = data_complete)
```

```{r}
tbl_regression(cc.mntl_rcs, 
               exponentiate = TRUE) %>%
  bold_p() %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  as_gt() %>%  # Convert to gt for further customization
  gt::tab_header(
    title = "Logistic Regression: TBI Aetiology by Mental Health Treatment"
  )
```

```{r}
cc.suicide_rcs <- glm(Aetiology ~ Suicide_Lifetime + rcs(Age, 3) + Sex, family = binomial, data = data_complete)
```

```{r}
tbl_regression(cc.suicide_rcs, 
               exponentiate = TRUE) %>%
  bold_p() %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  as_gt() %>%  # Convert to gt for further customization
  gt::tab_header(
    title = "Logistic Regression: TBI Aetiology by Suicide Attempt"
  )
```

```{r}
cc.drink_rcs <- glm(Aetiology ~ Alcohol_Use + rcs(Age, 3) + Sex, family = binomial, data = data_complete)
```

```{r}
tbl_regression(cc.drink_rcs, 
               exponentiate = TRUE) %>%
  bold_p() %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  as_gt() %>%  # Convert to gt for further customization
  gt::tab_header(
    title = "Logistic Regression: TBI Aetiology by Alcohol Use"
  )
```

```{r}
cc.drug_rcs <- glm(Aetiology ~ Drugs + rcs(Age, 3) + Sex, family = binomial, data = data_complete)

```

```{r}
tbl_regression(cc.drug_rcs, 
               exponentiate = TRUE) %>%
  bold_p() %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  as_gt() %>%  # Convert to gt for further customization
  gt::tab_header(
    title = "Logistic Regression: TBI Aetiology by Illicit Drug Use"
  )
```


## Fully adjusted models 

Adjust for all sociodemographic risk factors (age, sex, race, urbanicity, educational attainment, learning difficulties/special education, and employment status) in the logistic regression models.

```{r}
cc.psych.full <- glm(Aetiology ~ Psychiatric_Hospitalisation + rcs(Age, 3) + Sex
                          + Race + Urbanicity + Education + Special_Education + Employment, family = binomial, data = data_complete)
```

```{r}
tbl_regression(cc.psych.full, 
               exponentiate = TRUE) %>%
  bold_p() %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  as_gt() %>%  # Convert to gt for further customization
  gt::tab_header(
    title = "Logistic Regression: TBI Aetiology by Psychiatric Hospitalisation (Fully Adjusted Model)"
  )
```


```{r}
cc.mntl.full <- glm(Aetiology ~ Mental_Treatment + rcs(Age, 3) + Sex
                           + Race + Urbanicity + Education + Special_Education + Employment, family = binomial, data = data_complete)
```

```{r}
tbl_regression(cc.mntl.full, 
               exponentiate = TRUE) %>%
  bold_p() %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  as_gt() %>%  # Convert to gt for further customization
  gt::tab_header(
    title = "Logistic Regression: TBI Aetiology by Mental Health Treatment (Fully Adjusted Model)"
  )
```

```{r}
cc.suicide.full <- glm(Aetiology ~ Suicide_Lifetime + rcs(Age, 3) + Sex
                           + Race + Urbanicity + Education + Special_Education + Employment, family = binomial, data = data_complete)
```

```{r}
tbl_regression(cc.suicide.full, 
               exponentiate = TRUE) %>%
  bold_p() %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  as_gt() %>%  # Convert to gt for further customization
  gt::tab_header(
    title = "Logistic Regression: TBI Aetiology by Suicide Attempt (Fully Adjusted Model)"
  )
```

```{r}
cc.drink.full <- glm(Aetiology ~ Alcohol_Use + rcs(Age, 3) + Sex
                           + Race + Urbanicity + Education + Special_Education + Employment, family = binomial, data = data_complete)
```

```{r}
tbl_regression(cc.drink.full, 
               exponentiate = TRUE) %>%
  bold_p() %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  as_gt() %>%  # Convert to gt for further customization
  gt::tab_header(
    title = "Logistic Regression: TBI Aetiology by Alcohol Use (Fully Adjusted Model)"
  )
```

```{r}
cc.drug.full <- glm(Aetiology ~ Drugs + rcs(Age, 3) + Sex
                           + Race + Urbanicity + Education + Special_Education + Employment, family = binomial, data = data_complete)
```

```{r}
tbl_regression(cc.drug.full, 
               exponentiate = TRUE) %>%
  bold_p() %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  as_gt() %>%  # Convert to gt for further customization
  gt::tab_header(
    title = "Logistic Regression: TBI Aetiology by Illicit Drug Use (Fully Adjusted Model)"
  )
```


# Clean up

Save data to the .RProject.

```{r}
save(object = longitudinal_complete, file = "data/longitudinal_complete.RData") # save the linked dataset
save(object = longitudinal_clean, file = "data/longitudinal_clean.RData") # cleaned data were I have recoded missing as NA. 
save(object = data_risk, file = "data/data_risk.RData") # data use for the psychiatric risk factor study
```

Clean environment.

```{r, results='hide'}
gc() # garabage, free unused R memory
rm(list = ls()) # remove all from R environment
cat("\014") # clear console
```
